Bennett Bishop work for Question 2:

2.  Temporarily remove the outlier trimming from preprocessing and do some exploratory analysis of outlying values. Are there specific *subjects* (not values) that seem to be outliers? If so, are outliers more frequent in one group or the other? (Hint: consider tabluating the number of outlying values per subject.)

``` {r}
## Preprocessing function but without trimming:

# get names
var_names <- read_csv('../../data/biomarker-raw.csv', 
                     col_names = F, 
                     n_max = 2, 
                     col_select = -(1:2)) %>%
  t() %>%
  as_tibble() %>%
  rename(name = V1, 
         abbreviation = V2) %>%
  na.omit()

# function for trimming outliers (good idea??)
# trim <- function(x, .at){
# x[abs(x) > .at] <- sign(x[abs(x) > .at])*.at
#  return(x)
#}

# read in data
biomarker_clean <- read_csv('../../data/biomarker-raw.csv', 
         skip = 2,
         col_select = -2L,
         col_names = c('group', 
                       'empty',
                       pull(var_names, abbreviation),
                       'ados'),
         na = c('-', '')) %>%
  filter(!is.na(group)) %>%
  # log transform, center and scale, and trim
  mutate(across(.cols = -c(group, ados),
                ~ (scale(log10(.x))[, 1]))) %>%
  # reorder columns
  select(group, ados, everything())

# calculate Z-scores for each column
z_score <- biomarker_clean %>%
  group_by(group) %>%
  mutate(across(where(is.numeric), ~ ((. - mean(., na.rm = TRUE)) / sd(., na.rm = TRUE)))) %>%
  ungroup()

# count outliers using `if_else()` and summarise per group
outlier_counts <- z_score %>%
  mutate(across(where(is.numeric), ~ if_else(abs(.) > 3, 1, 0))) %>%
  group_by(group) %>%
  summarise(across(where(is.numeric), sum, na.rm = TRUE))

# Display the outlier counts by group and variable
print(outlier_counts)
```

From simply looking over each subject above, it looks like most have 0-3 outliers. Lets see the total number of outliers in each group:

```{r}
total_outliers_by_group <- outlier_counts %>%
  rowwise() %>%
  mutate(Total_Outliers = sum(c_across(where(is.numeric)), na.rm = TRUE)) %>%
  select(group, Total_Outliers)

total_outliers_by_group
```
It looks like ASD has 1062 total outliers, while TD has a total of 1341 outliers. 

Now, we can check out the subjects with large numbers of outliers. The following graph shows subjects with 4+ outliers.

```{r}
# Identify subjects with 4 or more outliers
outliers_ge_4 <- outlier_counts %>%
  pivot_longer(cols = -group, names_to = "Column", values_to = "Outlier_Count") %>%
  filter(Outlier_Count >= 4) %>%
  arrange(group, desc(Outlier_Count))

# Display the result
outliers_ge_4
```

We can see that 4 groups above, groups: Sig14, TLR2, TGM3 and GM-CSF all have 4-5 outliers. Neat!
