---
title: "Untitled"
author: "Amy Ji"
date: "2024-10-29"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#problem 1 
Log-transforming protein values in a data set is done to address the distribution of the raw values. The data follows a right-skewed (positively skewed) distribution, meaning there are many small values and a few very large values. This skew can make it difficult to analyze patterns and correlations, as the larger values can overly influence results.

When you apply a log transformation, we would reduce the impact of the extreme high values, therefore making the distribution more symmetric and closer to a normal distribution. This action would bring outliers closer to the main body data, which would stabalize variance and improve the performance of models. Additionally, it could make the relationships between the variables more linear, which would ensure that in analyses the results will not be overly influenced by extreme values. 

#Problem 3 
```{r}
getwd()
setwd("/Users/amy/Desktop/pstat197A/module1-7")

library(tidyverse)
# get names
var_names <- read_csv("/Users/amy/Desktop/pstat197A/module1-7/data/biomarker-raw.csv", 
                     col_names = F, 
                     n_max = 2, 
                     col_select = -(1:2)) %>%
  t() %>%
  as_tibble() %>%
  rename(name = V1, 
         abbreviation = V2) %>%
  na.omit()
# function for trimming outliers (good idea??)
trim <- function(x, .at){
  x[abs(x) > .at] <- sign(x[abs(x) > .at])*.at
  return(x)
}

# read in data
biomarker_clean <- read_csv('data/biomarker-raw.csv', 
         skip = 2,
         col_select = -2L,
         col_names = c('group', 
                       'empty',
                       pull(var_names, abbreviation),
                       'ados'),
         na = c('-', '')) %>%
  filter(!is.na(group)) %>%
  # log transform, center and scale, and trim
  mutate(across(.cols = -c(group, ados), 
                ~ trim(scale(log10(.x))[, 1], .at = 3))) %>%
  # reorder columns
  select(group, ados, everything())

# export as r binary
save(list = 'biomarker_clean', 
     file = 'data/biomarker-clean.RData')
```

```{r}

library(tidyverse)
library(infer)
library(randomForest)
library(tidymodels)
library(modelr)
library(yardstick)


# partition data
biomarker_split <- biomarker_clean %>%
  initial_split(prop = 0.8)

biomarker_train <- training(biomarker_split)
biomarket_test <- testing(biomarker_split)

## MULTIPLE TESTING
####################

# function to compute tests
test_fn <- function(.df){
  t_test(.df, 
         formula = level ~ group,
         order = c('ASD', 'TD'),
         alternative = 'two-sided',
         var.equal = F)
}

ttests_out <- biomarker_train %>%
  # drop ADOS score
  select(-ados) %>%
  # arrange in long format
  pivot_longer(-group, 
               names_to = 'protein', 
               values_to = 'level') %>%
  # nest by protein
  nest(data = c(level, group)) %>% 
  # compute t tests
  mutate(ttest = map(data, test_fn)) %>%
  unnest(ttest) %>%
  # sort by p-value
  arrange(p_value) %>%
  # multiple testing correction
  mutate(m = n(),
         hm = log(m) + 1/(2*m) - digamma(1),
         rank = row_number(),
         p.adj = m*hm*p_value/rank)

# select significant proteins
proteins_s1 <- ttests_out %>%
  slice_min(p.adj, n = 20) %>%
  pull(protein)

## RANDOM FOREST
##################

# store predictors and response separately
predictors <- biomarker_train %>%
  select(-c(group, ados))

response <- biomarker_train %>% pull(group) %>% factor()

# fit RF
set.seed(101422)
rf_out <- randomForest(x = predictors, 
                       y = response, 
                       ntree = 1000, 
                       importance = T)

# check errors
rf_out$confusion

# compute importance scores
proteins_s2 <- rf_out$importance %>% 
  as_tibble() %>%
  mutate(protein = rownames(rf_out$importance)) %>%
  slice_max(MeanDecreaseGini, n = 20) %>%
  pull(protein)


## LOGISTIC REGRESSION

# Weighted Combination of Ranks (Fuzzy Intersection)

# Rank proteins by significance in each method
ttests_ranked <- ttests_out %>%
  mutate(ttest_rank = rank(p.adj)) %>%
  select(protein, ttest_rank)

rf_ranked <- rf_out$importance %>%
  as.data.frame() %>%
  rownames_to_column(var = "protein") %>%
  as_tibble() %>%
  mutate(rf_rank = rank(-MeanDecreaseGini)) %>%
  select(protein, rf_rank)

# Combine ranks and calculate combined score
combined_ranks <- ttests_ranked %>%
  inner_join(rf_ranked, by = "protein") %>%
  mutate(combined_score = ttest_rank + rf_rank) %>%
  arrange(combined_score)

# Select top proteins based on combined score
proteins_sstar <- combined_ranks %>%
  slice_min(combined_score, n = 20) %>%
  pull(protein)

biomarker_sstar <- biomarker_clean %>%
  select(group, any_of(proteins_sstar)) %>%
  mutate(class = (group == 'ASD')) %>%
  select(-group)

# partition into training and test set
set.seed(101422)
biomarker_split <- biomarker_sstar %>%
  initial_split(prop = 0.8)

# fit logistic regression model to training set
fit <- glm(class ~ ., 
           data = training(biomarker_split), 
           family = 'binomial')

# evaluate errors on test set
class_metrics <- metric_set(sensitivity, 
                            specificity, 
                            accuracy,
                            roc_auc)

testing(biomarker_split) %>%
  add_predictions(fit, type = 'response') %>%
  mutate(est = as.factor(pred > 0.5), tr_c = as.factor(class)) %>%
  class_metrics(estimate = est,
              truth = tr_c, pred,
              event_level = 'second')

```
```





